Deployable Probabilistic Programming

### 2019-08-11 paper/paper.tex
typo fixed
up-to-date list of library elementals
basic rather than library
changes in progress before camera ready

Let us illustrate the use of Infergo on a basic example serving
as a probabilistic programming equivalent of the "Hello, World!"
program --- inferring parameters of a unidimensional Normal
distribution. In this example, the model object holds the set of
observations from which the distribution parameters are to be
inferred:
\begin{lstlisting}
type ExampleModel struct {
  Data []float64
}
\end{lstlisting}

The \lstinline{Observe} method computes the log-likelihood of
the distribution parameters. The Normal distribution has two
parameters: mean $\mu$ and standard deviation $\sigma$. To ease
the inference, positive $\sigma$ is transformed to unrestricted
$\log \sigma$, and the parameter vector \lstinline{x} is $[\mu,
\log \sigma]$. \lstinline{Observe} first imposes a prior on the
parameters, and then conditions the posterior distribution on
the observations:

\begin{lstlisting}
// x[0] is the mean,
// x[1] is the log stddev of the distribution
func (m *ExampleModel)
     Observe(x []float64) float64 {
  // Our prior is a unit normal ...
  ll := Normal.Logps(0, 1, x...)
  // ... but the posterior is based 
  // on data observations.
  ll += Normal.Logps(x[0], math.Exp(x[1]),
                     m.Data...)
  return ll
}
\end{lstlisting}

@@

\begin{lstlisting}
hmc := &infer.HMC{
  Eps: 0.1,
}
samples := make(chan []float64)
hmc.Sample(m, x, samples)
for i := 0; i != 5000; i++ {
  x = <-samples
}
hmc.Stop()
\end{lstlisting}

@@

Functions are considered elementals (and must have a
registered derivative) if their signature is either of kind
\lstinline{func (float64, float64*) float64}, that is, 
one or more non-variadic \lstinline{float64} arguments and
\lstinline{float64} return value, or \lstinline{func([]float64) float64}.
For example, functions
\begin{itemize}
	\item \lstinline{func foo(float64, float64, float64) float64}
	\item \lstinline{func bar([]float64) float64}
\end{itemize}
are considered elemental, while functions
\begin{itemize}
  \item \lstinline{func fee(...float64) float64}
  \item \lstinline{func buz(int, float64) float64}
\end{itemize}
are not. Gradients for selected functions from the
\lstinline{math} package are pre-defined (\lstinline{Sqrt},
\lstinline{Exp}, \lstinline{Log}, \lstinline{Pow},
\lstinline{Sin}, \lstinline{Cos}, \lstinline{Tan}). Auxiliary
elemental functions with pre-defined gradients are provided in a
separate package.

Any user-defined function of appropriate kind can be used as an
elemental inside model methods. The gradient must be registered
for the function using a call to
\begin{lstlisting}
func RegisterElemental(
  f interface{}, 
  g func(value float64,
         params ...float64) []float64)
\end{lstlisting} For
example, the call
\begin{lstlisting}
RegisterElemental(Sigm,
  func(value float64,
       _ ...float64) []float64 {
    return []float64{value*(1-value)}
  })
\end{lstlisting}
registers the gradient for function
\lstinline{Sigm}$(x)\coloneq\left(1+e^{-x}\right)^{-1}$.

\subsection{Inference}

@@

\subsection{Model Composition}

@@

\clearpage
\clearpage

@@

\clearpage

>>>>

Let us illustrate the use of Infergo on a basic example serving
as a probabilistic programming equivalent of the "Hello, World!"
program --- inferring parameters of a unidimensional Normal
distribution given a set of data samples. The two parameters of
a Normal distribution are the mean $\mu$ and the standard
deviation $\sigma$. The solution can of course be obtained
analytically, but the organization of implementation and
inference possesses all of the elements of probabilistic
programming for larger, more complicated problems, for which a
closed form solution does not exist or difficult to derive.

In this example, the model object holds the set of observations
from which the distribution parameters are to be inferred:
\begin{lstlisting}
type ExampleModel struct {
  Data []float64
}
\end{lstlisting}

The \lstinline{Observe} method computes the log-likelihood of
the distribution parameters. The Normal distribution has two
parameters: mean $\mu$ and standard deviation $\sigma$. To ease
the inference, positive $\sigma$ is transformed to unrestricted
$\log \sigma$, and the parameter vector \lstinline{x} is $[\mu,
\log \sigma]$. \lstinline{Observe} first imposes a prior on the
parameters, and then conditions the posterior distribution on
the observations\footnote{A call to \lstinline{Normal.Logps}
computes the joint log-probability of observations from a Normal
distribution.}:

\begin{lstlisting}
// x[0] is the mean,
// x[1] is the log stddev of the distribution
func (m *ExampleModel)
     Observe(x []float64) float64 {
  // Our prior is a unit normal ...
  ll := Normal.Logps(0, 1, x...)
  // ... but the posterior is based on 
  // the data.
  ll += Normal.Logps(x[0], math.Exp(x[1]),
                     m.Data...)
  return ll
}
\end{lstlisting}

@@

\begin{lstlisting}
hmc := &infer.HMC{
  Eps: 0.1,
}
samples := make(chan []float64)
hmc.Sample(m, x, samples)
for i := 0; i != 5000; i++ {
  sample = <-samples
}
hmc.Stop()
\end{lstlisting}

@@

Commonly, elementals are basic mathematical functions.
Functions are considered elementals (and must have a
registered derivative) if their signature is either of kind
\lstinline{func (float64, float64*) float64}, that is,
one or more non-variadic \lstinline{float64} arguments and
\lstinline{float64} return value, or \lstinline{func([]float64) float64}.
For example, functions
\begin{itemize}
	\item \lstinline{func foo(float64, float64, float64) float64}
	\item \lstinline{func bar([]float64) float64}
\end{itemize}
are considered elementals, while functions
\begin{itemize}
  \item \lstinline{func fee(...float64) float64}
  \item \lstinline{func buz(int, float64) float64}
\end{itemize}
are not. Gradients for selected functions from the
\lstinline{math} package are pre-defined (\lstinline{Sqrt},
\lstinline{Exp}, \lstinline{Log}, \lstinline{Pow},
\lstinline{Sin}, \lstinline{Cos}, \lstinline{Tan},
\lstinline{Erf}, \lstinline{Erfc}). Auxiliary elemental
functions with pre-defined gradients are provided in a separate
package.

Any user-defined function of appropriate kind can be used as an
elemental inside model methods. The gradient must be registered
for the function using a call to
\begin{lstlisting}
func RegisterElemental(
  f interface{},
  g func(value float64,
         params ...float64) []float64)
\end{lstlisting} For
example, the call
\begin{lstlisting}
RegisterElemental(Sigm,
  func(value float64,
       _ ...float64) []float64 {
    return []float64{value*(1-value)}
  })
\end{lstlisting}
registers the gradient for the \textit{sigmoid} function:
\begin{lstlisting}
func Sigm(x float64) float64 {
	return 1. / (1. + math.Exp(-x))
}
\end{lstlisting}

\subsection{Inference}
\label{sec:inference}

@@

\subsection{Model Composition}
\label{sec:model-composition}

@@

\section*{Acknowledgements}

The author thanks PUB+ (\url{http://pubplus.com/}) for
supporting development of Infergo and application of
probabilistic programming to critical decision-making in
production enviroment.

@@
